Using HbA1c and Fasting BG to Predict Development of CFRD

### 2025-10-15 Christine Chan/CF Registry/cf_registry_CFRD_prediction.qmd
HbA1c joint model NACFC poster

```{r setup}
#| include: false
library(survival)
library(nlme)
library(JM)
library(dplyr)
library(tidyr)
library(arsenal)
library(knitr)
home_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

  demo %>% select(eDWID, Race, Hispanicrace, Gender, F508, MutClass),

@@

  select(

@@

  select(eDWID, ReviewYear, Age_YrEnd, all_of(predictors), cfrd_registry) %>%

@@

```{r}
# Formulas
cox_f <- as.formula(paste0(
  "Surv(time = Age_YrEnd, event = cfrd_registry)~",
  paste0(predictors, collapse = "+")
))
lme_f <- as.formula(paste0(
  "A_hgba1c~Age_YrEnd+",
  paste0(predictors, collapse = "+")
))
# Fit Cox model first, and then pull the clusters
cox_fit <- coxph(
  cox_f,
  cluster = eDWID,
  x = T,
  model = T,
  data = surv_reg[surv_reg$eDWID %in% reg_include, ]
)
lme_fit_a1c <- lme(
  lme_f,
  random = ~ Age_YrEnd | eDWID,
  na.action = na.omit,
  data = registry[registry$eDWID %in% reg_include, ]
)
# Combine
joint_fit_value <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd")
# joint_fit_slope <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd",parameterization ="slope")
# joint_fit_both <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd",parameterization ="both")
s <- summary(joint_fit_value)
long <- data.frame(s$`CoefTable-Long`)
long$p.value <- format.pval(long$p.value, eps = 0.001, digits = 3)
event <- data.frame(s$`CoefTable-Event`)
event$Value <- exp(event$Value)
rownames(event)[nrow(event)] <- "shape"
event$p.value <- format.pval(event$p.value, eps = 0.001, digits = 3)
# summary(joint_fit_slope)
# summary(joint_fit_both)
kable(long, caption = "Longitudinal Process", digits = 3)
kable(event, caption = "Event Process", digits = 3)
```

```{r}
nd <- joint_fit_value$data
nd$pred <- predict(
  joint_fit_value,
  registry[registry$eDWID %in% reg_include, ],
  type = "Subject",
  idVar = "eDWID"
)
```

>>>>

```{r setup}
#| include: false
library(tidyverse)
library(survival)
library(nlme)
library(JM)
library(arsenal)
library(knitr)
library(ggforestplot)
library(patchwork)
home_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

  demo %>% dplyr::select(eDWID, Race, Hispanicrace, Gender, F508, MutClass),

@@

  dplyr::select(

@@

  dplyr::select(

@@

```{r}
# Formulas
cox_f <- as.formula(paste0(
  "Surv(time = Age_YrEnd, event = cfrd_registry)~",
  paste0(predictors, collapse = "+")
))
lme_f <- as.formula(paste0(
  "A_hgba1c~Age_YrEnd+",
  paste0(predictors, collapse = "+")
))
# Fit Cox model first, and then pull the clusters
cox_fit <- coxph(
  cox_f,
  cluster = eDWID,
  x = T,
  model = T,
  data = surv_reg[surv_reg$eDWID %in% reg_include, ]
)
lme_fit_a1c <- lme(
  lme_f,
  random = ~ Age_YrEnd | eDWID,
  na.action = na.omit,
  data = registry[registry$eDWID %in% reg_include, ]
)
# Combine
joint_fit_value <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd")
# joint_fit_slope <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd",parameterization ="slope")
# joint_fit_both <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd",parameterization ="both")
s <- summary(joint_fit_value)
long <- data.frame(s$`CoefTable-Long`)
long$p.value <- format.pval(long$p.value, eps = 0.001, digits = 3)
event <- data.frame(s$`CoefTable-Event`)
event$Value <- exp(event$Value)
rownames(event)[nrow(event)] <- "shape"
event$p.value <- format.pval(event$p.value, eps = 0.001, digits = 3)
# summary(joint_fit_slope)
# summary(joint_fit_both)
kable(long, caption = "Longitudinal Process", digits = 3)
kable(event, caption = "Event Process", digits = 3)

# Plots
plot_df = lme_fit_a1c$data
plot_df$pred = predict(lme_fit_a1c)
p1 = ggplot(plot_df, aes(x = Age_YrEnd, y = pred)) +
  geom_point(alpha = 0.1, shape = ".") +
  geom_abline(intercept = 5.61, slope = 0.005895, color = "#3BA3BD") +
  ylab("HbA1c (%)") +
  xlab("Age (years)") +
  ylim(c(2.5, 15)) +
  ggtitle("Longitudinal Process") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

fp_df = data.frame(s$`CoefTable-Event`) %>%
  rownames_to_column() %>%
  filter(!rowname %in% c("(Intercept)", "log(shape)")) %>%
  rename(estimate = "Value", term = "rowname") %>%
  mutate(
    conf.low = estimate - (1.96 * Std.Err),
    conf.high = estimate + (1.96 * Std.Err)
  )
fp_df$term = factor(
  fp_df$term,
  levels = c(
    "GenderF",
    "RaceBlack or African American",
    "RaceOther race",
    "RaceMore than one race",
    "HispanicraceNo",
    "A_IsOnEnzymesNo",
    "NumPulmExacerbation1",
    "NumPulmExacerbation2",
    "NumPulmExacerbation>=3",
    "A_FEV1_pct_predicted",
    "bmi",
    "Assoct"
  ),
  labels = c(
    "Female",
    "Black or African American",
    "Other race",
    "More than one race",
    "Hispanic",
    "Pancreatic sufficient",
    "1 PEx in past year",
    "2 PEx in past year",
    ">= 3 PEx in past year",
    "FEV1 % predicted",
    "BMI",
    "Association Parameter*"
  )
)
fp_df$term = as.character(fp_df$term)
fp_df$variable = fp_df$p.value <= 0.05
fp_df$estimate = exp(fp_df$estimate)
fp_df$conf.low = exp(fp_df$conf.low)
fp_df$conf.high = exp(fp_df$conf.high)
p2 = forestplot(
  df = fp_df,
  name = term,
  estimate = estimate,
  se = Std.Err,
  xlab = "Hazard ratio for CFRD",
  title = "Event Process",
  colour = variable
) +
  scale_color_manual(values = c("black", "#3BA3BD")) +
  theme(legend.position = "none")
p1 + p2
```
